Validation of Visual Statistical Inference, Applied to Linear Models

### 2012-07-04 paper_jasa.tex
uncontaminated p-value added
some typo fixed

%\red{pull the results for $P(p_D < p_{i,0})$ and $P(p_D < p_0)$ to the front, including the figure on $1/m$ versus power. These results are independent of the regression setting - they don't make any assumptions in terms of regression.}

Figure \ref{fig:pval_power} gives an overview of the probability of picking the actual data plot for lineups of different size: as $m$ increases we have an increased probability to observe a more highly structured null plot by chance. We see that for a $p$-value $p_D$ of about 0.15 for the data plot the signal in the plot is so weak, that we cannot distinguish the actual data plot from null plots in a lineup of size $m=20$. % This pattern is confirmed by our findings described in Section~\ref{sec:simulation}.
%also seen in our experiments as shown in figure \ref{fig:pval_pcorrect}.

@@

The number of evaluations required for each lineup to provide reasonable estimates of the power is determined by the variance of the estimated power. Suppose $\gamma$ denotes the conventional test power for each parameter combination shown in table \ref{tbl:experiment_params}. Since the expected power of visual inference is very close to the power of conventional test (Figure \ref{fig:power_expected}) the proportion of successful evaluations of lineups should give a rough estimate for  $\gamma$. For a given proportion $\gamma$ it is desired to have a margin of error (ME) less than or equal to 0.05. Thus we have $ME =1.96 \sqrt{ \frac 1 n_{\gamma} \gamma(1-\gamma)} \le 0.05$ which gives us the estimation of minimum number of evaluations $$n_{\gamma} \geq \frac{\gamma(1-\gamma)}{(0.05/1.96)^2}.$$

@@

%For a given proportion $p$ we want to have margin of error (ME) to be 0.1. Thus we have $ME =1.96 \sqrt{ \frac 1 n p(1-p)} \le 0.05$ which gives us the estimation of minimum number of evaluations $$n_p \geq \frac{p(1-p)}{(0.05/1.96)^2}.$$ Now from the theoretical (UMP) power curve for each parameter combinations shown in table \ref{tbl:experiment_params} we know the power(say $p$) and thus we can estimate the number of evaluations required for that specific combinations of parameter. We collect data such that each of the parameter combinations in the table \ref{tbl:experiment_params} has approximately the estimated sample size that we calculated here.

@@

%{\bf Presenting lineups for evaluation} Each subject got at least 10 lineups for evaluation. For each combination of sample size ($n$) and error SD ($\sigma$) we define the difficulty of the lineup based on the five slope ($\beta$) parameters we choose as shown in table \ref{tbl:experiment_params}. For the smallest slope parameter the difficulty is 0 and for the largest parameter the difficulty is 4. The 10 lineups are chosen as follows;

@@

%\begin{figure*}[hbtp]
%   \centering
%       \scalebox{0.7}{\includegraphics{power_diff_exp.pdf}}
%       \caption{Difference of Observed power from UMP test power obtained in three experiments for sample size $n= 100,300$ and $\sigma = 12,5$ and values of slope parameter $\beta$ as shown in table \ref{tbl:experiment_params}.}
%       \label{fig:power_diff_exp}
%\end{figure*}

@@

\begin{figure*}[hbtp]
   \centering
       \scalebox{0.6}{\includegraphics{p_val_prop_correct.pdf}}
       \caption{Proportion of correct responses decreases rapidly with the increase of p-value. After p-value exceeds 0.15 it is very unlikely to identify the actual plot. The theoretical justification of this is shown in figure \ref{fig:pval_power}. }
       \label{fig:pval_pcorrect}
\end{figure*}

@@

\begin{figure*}[hbtp]
   \centering
       \scalebox{0.6}{\includegraphics{p_val_definition.pdf}}
       \caption{Conventional test $p$-value ($p_D$) vs visual $p$-value obtained from the definition . p-values are shown on square root scale scale. }
    %   \hh{Mahbub, another change: could you change the y label to 'estimate of visual p-value $\hat{p_D}$ ?}
       \label{fig:pval_definition}
\end{figure*}

@@

%Number of lineups for which most of the subjects picked the data plot based on the minimum $p$-values are shown in table \ref{tbl:pval_assumption} for both experiment 1 and 2. The $p$-values of lineups which were not chosen based on minimum $p$-value are also very small (close to minimum) as we see in both figures \ref{fig:P-val_log} and \ref{fig:P-val_log2}. We see the same pattern in figure \ref{fig:P-val_rank} where the median number of people who pick the data plot with minimum $p$-value (rank 1) are much higher than other ranks.

@@

\noindent It was possible to choose more than one reason. The question begged for ``visible trend'' (2) to be the reasons that observers gave for correctly choosing the actual data plot. Either reasons 1 or 3 would be indicative of Type III error. It is interesting to calculate the proportion of observers who choose the correct plot, but use reasons 1 or 3, relative to reason 2. Figure \ref{fig:choice_reason} examines this. For observers who chose the actual data plot, the proportions for the three reasons are shown. It can be seen that the most common reason was ``most different plot'' (1), followed by clustering visible (3), with ``visible trend'' coming in third. \green{What about other??} \blue{Not many people chose other as we see in figure \ref{fig:choice_reason}.}

@@

\green{I find this plot really confusing! Can we just look at whether they selected choice ``clustering visible'' or not, if they were correct?} \blue{Is the right hand side plot in figure \ref{fig:choice_reason} ok? Should we remove the left plot?}

>>>>

%\red{pull the results for $P(p_D < p_{i,0})$ and $P(p_D < p_0)$ to the front, including the Figure on $1/m$ versus power. These results are independent of the regression setting - they don't make any assumptions in terms of regression.}

Figure \ref{fig:pval_power} gives an overview of the probability of picking the actual data plot for lineups of different size: as $m$ increases we have an increased probability to observe a more highly structured null plot by chance. We see that for a $p$-value $p_D$ of about 0.15 for the data plot the signal in the plot is so weak, that we cannot distinguish the actual data plot from null plots in a lineup of size $m=20$. % This pattern is confirmed by our findings described in Section~\ref{sec:simulation}.
%also seen in our experiments as shown in Figure \ref{fig:pval_pcorrect}.

@@

The number of evaluations required for each lineup to provide reasonable estimates of the power is determined by the variance of the estimated power. Suppose $\gamma$ denotes the conventional test power for each parameter combination shown in Table \ref{tbl:experiment_params}. Since the expected power of visual inference is very close to the power of conventional test (Figure \ref{fig:power_expected}) the proportion of successful evaluations of lineups should give a rough estimate for  $\gamma$. For a given proportion $\gamma$ it is desired to have a margin of error (ME) less than or equal to 0.05. Thus we have $ME =1.96 \sqrt{ \frac 1 n_{\gamma} \gamma(1-\gamma)} \le 0.05$ which gives us the estimation of minimum number of evaluations $$n_{\gamma} \geq \frac{\gamma(1-\gamma)}{(0.05/1.96)^2}.$$

@@

%For a given proportion $p$ we want to have margin of error (ME) to be 0.1. Thus we have $ME =1.96 \sqrt{ \frac 1 n p(1-p)} \le 0.05$ which gives us the estimation of minimum number of evaluations $$n_p \geq \frac{p(1-p)}{(0.05/1.96)^2}.$$ Now from the theoretical (UMP) power curve for each parameter combinations shown in Table \ref{tbl:experiment_params} we know the power(say $p$) and thus we can estimate the number of evaluations required for that specific combinations of parameter. We collect data such that each of the parameter combinations in the Table \ref{tbl:experiment_params} has approximately the estimated sample size that we calculated here.

@@

%{\bf Presenting lineups for evaluation} Each subject got at least 10 lineups for evaluation. For each combination of sample size ($n$) and error SD ($\sigma$) we define the difficulty of the lineup based on the five slope ($\beta$) parameters we choose as shown in Table \ref{tbl:experiment_params}. For the smallest slope parameter the difficulty is 0 and for the largest parameter the difficulty is 4. The 10 lineups are chosen as follows;

@@

%\begin{figure*}[hbtp]
%   \centering
%       \scalebox{0.7}{\includegraphics{power_diff_exp.pdf}}
%       \caption{Difference of Observed power from UMP test power obtained in three experiments for sample size $n= 100,300$ and $\sigma = 12,5$ and values of slope parameter $\beta$ as shown in Table \ref{tbl:experiment_params}.}
%       \label{fig:power_diff_exp}
%\end{figure*}

@@

\begin{figure*}[hbtp]
   \centering
       \scalebox{0.6}{\includegraphics{p_val_prop_correct.pdf}}
       \caption{Proportion of correct responses decreases rapidly with the increase of p-value. After p-value exceeds 0.15 it is very unlikely to identify the actual plot. The theoretical justification of this is shown in Figure \ref{fig:pval_power}. }
       \label{fig:pval_pcorrect}
\end{figure*}

@@

\begin{figure*}[hbtp]
   \centering
       \scalebox{0.55}{\includegraphics{p_val_definition.pdf}}
       \caption{Conventional test $p$-value ($p_D$) vs visual $p$-value obtained from the definition . Values are shown on square root scale. For experiment 3, $p$-values of uncontaminated part of the data are also shown.}
    %   \hh{Mahbub, another change: could you change the y label to 'estimate of visual p-value $\hat{p_D}$ ?}
       \label{fig:pval_definition}
\end{figure*}

@@

%Number of lineups for which most of the subjects picked the data plot based on the minimum $p$-values are shown in Table \ref{tbl:pval_assumption} for both experiment 1 and 2. The $p$-values of lineups which were not chosen based on minimum $p$-value are also very small (close to minimum) as we see in both Figures \ref{fig:P-val_log} and \ref{fig:P-val_log2}. We see the same pattern in Figure \ref{fig:P-val_rank} where the median number of people who pick the data plot with minimum $p$-value (rank 1) are much higher than other ranks.

@@

\noindent It was possible to choose more than one reason. The question begged for ``visible trend'' (2) to be the reasons that observers gave for correctly choosing the actual data plot. Either reasons 1 or 3 would be indicative of Type III error. It is interesting to calculate the proportion of observers who choose the correct plot, but use reasons 1 or 3, relative to reason 2. Figure \ref{fig:choice_reason} examines this. For observers who chose the actual data plot, the proportions for the three reasons are shown. It can be seen that the most common reason was ``most different plot'' (1), followed by clustering visible (3), with ``visible trend'' coming in third. \green{What about other??} \blue{Not many people chose other as we see in Figure \ref{fig:choice_reason}.}

@@

\green{I find this plot really confusing! Can we just look at whether they selected choice ``clustering visible'' or not, if they were correct?} \blue{Is the right hand side plot in Figure \ref{fig:choice_reason} ok? Should we remove the left plot?}
